---
title: "Data 206 Capstone EDA Project"
format: ipynb
---

## Installing libraries

```{r}
library(dplyr)
library(ggplot2)
library(lubridate)
```

```{r}
Impound <- read.csv("C:/Users/casti/OneDrive/Documents/Data 206 Capstone Project/.quarto/project-cache/Animal Impound.csv")
Shelter <- read.csv("C:/Users/casti/OneDrive/Documents/Data 206 Capstone Project/.quarto/project-cache/Animal Pathway.csv") #Adding the two datasets


```

```{r}

colnames(Impound)[colnames(Impound) == "Animal.identification.number"] <- "Animal.ID"
colnames(Shelter)[colnames(Shelter) == "Animal.identification.number"] <- "Animal.ID" # Changing the name of the column to make it easier
combined <- left_join(Impound,Shelter, by = "Animal.ID") #The merging of the two datasets
sum(Impound$Animal.ID %in% Shelter$Animal.ID) # To see how many of the ID joined
combined$Departure.date.and.time[combined$Departure.date.and.time == ""] <- NA # To remove empty space
names(Shelter) # To see some of the variables for further exploration

```

```{r, include=FALSE}
combined <- combined %>%
  rename(
    Age_Year = Pet.s.age.in.years,
    Animal_Type = Animal.type,
    Bite_History = Bite.history,
    Color = Color.of.animal,
    Breed = Breed.of.animal
  )
```

```{r, include=FALSE}
combined <- combined %>%
  mutate(
    Departure = mdy_hm(Departure.date.and.time),
    Intake = mdy_hm(Impound.date.and.time), #Creating new variables for datetime to take the difference.
    LengthOfStay = round(as.numeric(difftime(Departure, Intake, units = "days")), 2)
  )
```

```{r, include=FALSE}

combined <- combined %>%
  mutate(
    StayDays = floor(LengthOfStay),#Looking at the first number of the decimal for the hour 
    

    StayHours = floor((LengthOfStay - floor(LengthOfStay)) * 24), # Removes the front decimal and looks at the second digit to get hours
    
    StayMinutes = round(((LengthOfStay - floor(LengthOfStay)) * 24 - StayHours) * 60) # Looks at the last digit of the decimal and creates it into minutes
  )

```

```{r, include=FALSE}
Year_month <- combined %>%
  filter(!is.na(Departure.date.and.time)) # Made a new dataframe for Yearmonth due to complications of doing it to combined dataframe and filter for no NAs

Year_month$Departure.date.and.time[Year_month$Departure.date.and.time == ""] <- NA # Getting rid of negative space as some inputs in the row were " " to the word NA
Year_month <- combined %>%
  filter(!is.na(Departure.date.and.time)) # Had to do it twice as the spaces were now turned to NAs
head(Year_month$Departure.date.and.time) # To check for NAs
```

```{r,include=FALSE}
Year_month$Departure.date.and.time <- as.POSIXct(
 Year_month$Departure.date.and.time,
  format = "%m/%d/%Y %H:%M" # Had to do this again different method to get the format right this time
)
Year_month$YearMonth <- format(Year_month$Departure.date.and.time, "%Y-%m")
# Having year/month as a new variable to be part of my graph below to have a compact x axis
```

```{r}
monthly_outcomes <- Year_month %>%
  filter(!is.na(Departure.date.and.time)) %>% # making sure there is no NAs 
  group_by(YearMonth, Outcome.type.y) %>% # To create a group for each month for each outcome
  summarise(count = n(), .groups = "drop") #To make each outcome a new row for each month
```

\

# These graphs could possibly supplement my tableau map

## First Graph for Project - Are certain times of the year associated with higher intake rates?​

```{r}
ggplot(monthly_outcomes, aes(x = YearMonth, y = count, fill = Outcome.type.y)) +
  geom_bar(stat = "identity") + # Make sure that it uses the counts from montly income
  theme_minimal() + # Add theme to my graph
   geom_text(aes(label = count),
            position = position_stack(vjust = 0.5), # This will center my count in the middle of each stack
            size = 3,
            color = "black") +
  scale_fill_brewer(palette = "Set1")  + 
  labs(
    title = "Monthly Shelter Outcomes",
    x = "Year/Month of Year 2025 to February 2026",
    y = "Number of Animals",
    fill = "Outcome Type"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) # This makes it possible so the x labels don't overlap

```

## Creating second data frame for my next graph

```{r}
Euthanized <- combined %>%
  filter(Outcome.type.y == "EUTH") #Filter for Euthanization

Euthanized <- Euthanized %>%
  filter(!is.na(Age_Year)) #Remove any NAs
Euthanized <- Euthanized %>%
  mutate(
    Age_group = case_when(
     Age_Year < 1 ~ "Under 1", # Create age groups as every number will be hard to graph
      Age_Year >= 1 & Age_Year < 3 ~ "1-3 Age group",
      Age_Year >= 3 & Age_Year < 7 ~ "3-7 Age group",
      Age_Year >= 7& Age_Year < 10 ~ "7-10 Age group",
      Age_Year > 10 ~ "10+ Age group"
    )
  )
Euthanized <- Euthanized %>%
  filter(!is.na(Age_group), !is.na(Animal_Type)) # Make sure any Nas are gone

Euthanized$Age_group <- factor(Euthanized$Age_group, levels = c("Under 1","1-3 Age group", "3-7 Age group","7-10 Age group","10+ Age group" )) # To change the order of the levels in which they present in the barchart below
```

## Second Graph- Does age affect euthanizing likelihood?​

```{r}
ggplot(Euthanized, aes(x = Age_group, fill = Animal_Type)) +
  geom_bar(position = "dodge") +
  theme_minimal() +
   geom_text(
    stat = "count",
    aes(label = ..count..),
    position = position_dodge(width = 0.9), #Helps align the number in the middle of each bar
    vjust = -0.3, #Helps align the number in the middle of each bar
    size = 3
  ) +
  labs(
    title = "Animal Type by Age Group For all Euthanizations",
    x = "Age Group",
    y = "Count of Animals",
    fill = "Animal Type"
  )
```

## Creating a new dataframe to get all the animal outcomes in percentage

```{r}

summary_data <- combined %>%
  filter(!is.na(Outcome.type.y)) %>%       # remove missing outcomes
  group_by(Outcome.type.y) %>%
  summarise(count = n()) %>%
  mutate(percentage = count / sum(count) * 100)

head(summary_data)
```

## Creating the third graph- What percentage of animals are adopted vs. euthanized vs. transferred?​

```{r}

ggplot(summary_data, aes(x = Outcome.type.y, y = percentage, fill = Outcome.type.y)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = paste0(round(percentage,1), "%")), vjust = -0.5) +
  labs(title = "Animal Outcomes", y = "Percentage (%)", x = "", fill = "Outcomes") +
  theme_minimal()

```

## Creating a dataframe for the different breeds of animals to remove any NAs

```{r}

breed_counts <- combined %>%
  filter(!is.na(Breed)) %>%       
  group_by(Breed) %>%
  summarise(count = n()) %>%
  arrange(desc(count))        

```

## Creating the Fourth Graph- What are the most common breeds of animals entering the shelter system?​

```{r}


top_breeds <- breed_counts %>%
  slice_max(count, n = 10)     # pick top 10


ggplot(top_breeds, aes(x = Breed, y = count, fill = Breed)) +
  geom_bar(stat = "identity") +
  coord_flip() +                     
  labs(title = "Top 10 Most Frequently Impounded Breeds",
       x = "Breed",
       y = "Number of Impounds") +
  theme_minimal() +
  theme(legend.position = "none")    # hide legend 

```

## Just wanted to see the mean number of days that the animals are at the shelter

```{r}
mean(combined$StayDays, na.rm = TRUE)
mean(combined$Age_Year, na.rm = TRUE)
```

\

## Creating the bite dataframe to remove NAs

```{r}
Bite_animal <- combined %>%
  filter(!is.na(Bite_History), Bite_History !="") #Removing NAs and blanks
```

\
\

## Creating fifth graph- Does the animal bite history affect its outcome?​

```{r}
ggplot(Bite_animal %>% filter(Bite_History == "N"),
                 aes(x = Outcome.type.y, fill = Outcome.type.y)) +
  
  geom_bar() + labs(title = "No Bite History", x = "Animal Outcome") + theme_minimal() + theme(legend.position = "none") 

 ggplot(Bite_animal %>% filter(Bite_History == "Y"),
                 aes(x = Outcome.type.y, fill = Outcome.type.y)) +
   
  geom_bar() + labs(title = "With Bite History",x = "Animal Outcome") + theme_minimal() + theme(legend.position = "none")

```

## Creating the Sixth Graph- How long do animals typically stay in the shelter?

```{r}
  Year_month %>%  
  group_by(YearMonth) %>%
  summarise(avg_stay = mean(StayDays, na.rm = TRUE)) %>%
  
  ggplot(aes(x = YearMonth, y = avg_stay, group = 1)) + # Creates a continous line
  geom_line(size = 1.2) +
  geom_point() +
  labs(
    title = "Average Shelter Stay by Month",
    x = "Year-Month",
    y = "Average Stay (Days)"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
combined$Adopted <- ifelse(combined$Outcome.type.y == "ADOPTION", 1, 0) ##STar fix
breed_adopt <- combined %>%
  group_by(Breed) %>%
  summarise(
    adoption_rate = mean(Adopted, na.rm = TRUE),
    count = n()
  ) %>%
   filter(count >= 20) %>%
  arrange(desc(adoption_rate)) %>%
  slice_head(n = 10)
breed_adopt$adoption_rate <- as.numeric(breed_adopt$adoption_rate)
```

```{r}
ggplot(breed_adopt, aes(x = reorder(Breed, adoption_rate), y = adoption_rate, fill = Breed)) +
  geom_col() +
  coord_flip() +
  scale_fill_brewer(palette = "Set3") +
  labs(title = "Top 10 Breeds by Adoption Rate",
       x = "Breed",
       y = "Adoption Rate") +
  scale_y_continuous(labels = scales::percent) +
   theme(legend.position = "none")
```

```{r}
euth_data <- combined %>%
  mutate(Euthanasia = Outcome.type.y == "EUTH") %>%
  group_by(Intake.type) %>%
  summarise(
    total = n(),
    euth_count = sum(Euthanasia, na.rm = TRUE),
    euth_rate = euth_count / total
  )
```

```{r}
ggplot(euth_data, aes(x = reorder(Intake.type, euth_rate), y = euth_rate)) +
  geom_col(fill = "red") +
  coord_flip() +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Euthanasia Rate by Intake Type",
       x = "Intake Type",
       y = "Euthanasia Rate") +
  theme_minimal()
```
